---
title: "ICT_LIS_661_Project_3"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to Project 3!

Before moving on, please make a copy of this file and rename it in a way that makes a clear distinction between the original file and the file you'll eventually submit.

> Under the headers below, I'll provide you with certain instructions, all of them formatted as blockquotes, like this paragraph is. Under each set of instructions, leave a blank line and then type your (non-blockquoted) response on the line after the blank one. At the very end of this document, I'll provide instructions for *knitting* and submitting the `.Rmd` to Canvas.

## Project 3, Part 1: Your Dataset [1 point]

> For this project, you will need to identify a dataset with at least 100 observations and 5 variables (at least 3 of which are quantitative). A dataset with at least 1,000 observations and 10 variables would be even better—but not necessary. This does not need to be a publicly available dataset, but please ensure that you can use the dataset legally and ethically. Store the data somewhere in your GitHub repository for this class.

> It is fine for this to be a dataset that you have used previously in this class; in fact, *Part 1* of this project repeats much of Project 2. There's real value in repeating that experience on new data, but I also won't blame you for saving yourself some time and copying over text and code from Project 2. However, if you are reusing a dataset, you must make sure that the dataset is appropriate for performing the analyses I am asking you to do here!

> In the code chunk below, please load the dataset into this document. (Keep in mind that you can—and may need to—load packages as part of this or any other code chunk; however, please do not try to install any packages here). [.5 points]

```{r}
Salary_Data <- read.csv("Salary_Data.csv", check.names = FALSE)

head(Salary_Data)
```


> As needed, perform any data cleaning (including "tidying" or "wrangling") on the data in the code chunk below. [.5 points]

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(infer)
library(moderndive)
```

## Project 3, Part 2: Exploratory Data Analysis [3 points]

> In Modules 9 and 10, we emphasized that it is important to explore our data before we proceed to analysis. For this part of the project, you will therefore carry out some exploratory data analysis.

### Examine Raw Data

> In the code chunk below, run some code for examining (some of) the raw data in your dataset. [1 point]

```{r}
summary_salary <- Salary_Data |>
group_by(`Education Level`, Gender) |>
summarize(
  mean_salary   = round(mean(Salary, na.rm = TRUE), 2),
  median_salary = round(median(Salary, na.rm = TRUE), 2),
  n             = n(),
  .groups = "drop"
)
summary_salary

```

### Summarize Data

> In the code chunk below, run some code for calculating summary statistics for your dataset. [1 point]

```{r}
summary_salary <- Salary_Data |>
group_by(`Education Level`, Gender) |>
summarize(
  mean_salary   = round(mean(Salary, na.rm = TRUE), 2),
  median_salary = round(median(Salary, na.rm = TRUE), 2),
  sd_salary     = round(sd(Salary, na.rm = TRUE), 2),
  min_salary    = min(Salary, na.rm = TRUE),
  max_salary    = max(Salary, na.rm = TRUE),
  n             = n(),
  .groups = "drop"
)
summary_salary
```

### Visualize Data

> In the code chunk below, run some code for carrying out an appropriate visualization for at least one key variable in your dataset. [1 point]

```{r}

plot_data <- summary_salary |>
filter(!is.na(mean_salary))

ggplot(plot_data, aes(x = `Education Level`, y = mean_salary, fill = Gender)) +
  geom_col(position = "dodge") +
  labs(
 title = "Average Salary by Education Level and Gender",
    x = "Education Level",
    y = "Average Salary ($)"
  ) +
  theme_minimal() +
  theme(
  plot.title = element_text(hjust = 0.5, face = "bold"),  
 axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Project 3, Part 2: Comparing Means or Proportions [3 points]

> In Module 13, we learned how to use hypothesis testing to compare two means or proportions. For this part of the project, you will be carrying out a hypothesis test for a similar comparison involving your own data.

### Hypotheses

> Below, please describe a null hypothesis and an alternative hypothesis related to your data that involves comparing means or proportions. [1 point]

The null hypothesis I have created for my dataset is: There is no significant relationship between both Years of Experience and Education level, and the salary someone earns. The alternative hypothesis I have created for my dataset is: There is a significant relationship between both Years of Experience and Education level, and the salary someone earns. 


### Data Analysis

> In the code chunk below, please carry out a test of the hypotheses you described above. Then, below the code chunk, summarize the results of this analysis; pay particular attention to statistical significance! [1 points]

```{r}
ggplot(Salary_Data, aes(x = Age, y = Salary)) +
  geom_point(color = "royalblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Relationship Between Age and Salary",
    x = "Age",
    y = "Salary ($)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 10)
  )

ggplot(Salary_Data, aes(x = `Years of Experience`, y = Salary, color = `Education Level`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
labs(x = "Age", y = "Salary", color = "Education Group", title = "Relationship Between Years of experience and Salary", ) +
  geom_parallel_slopes(se = FALSE)
```


### Conclusions

> Below, interpret the results of this analysis (that is, describe what the results mean for your data and your hypotheses). Please also discuss any practical implications of these results (that is, explain how people might wish to think or act differently based on these results). [1 point]

From this data, what we can see is that there is a correlation between both years of experience and Education Level. Though for higher education, that being college, the advantage of having a higher education level starts to soften and the influence of the amount of experience starts to increase. This means the alternative hypothesis is true. 


## Project 3, Part 3: Regression [3 points]

> In Module 14, we learned how to apply hypothesis testing to a regression analysis (which explores the relationship between two variables. For this part of the project, you will be carrying out a hypothesis test for a regression involving your own data.

### Hypotheses

> Below, please describe a null hypothesis and an alternative hypothesis related to your data that can be tested using regression. [1 point]

The null hypothesis I have created for my dataset is: There is no significant relationship between both Age and Education level, and the salary someone earns. The alternative hypothesis I have created for my dataset is: There is a significant relationship between both Age and Education level, and the salary someone earns. 

### Data Analysis

> In the code chunk below, please carry out a regression that will test the hypotheses listed above. Then, below the code chunk, summarize the results of this analysis. Make particular reference to relevant statistics, including coefficient and p-value. [1 point]

```{r}
ggplot(data = Salary_Data, aes(x = `Education Level`, y = Salary)) +
  geom_boxplot() +
  labs(y = "Salary", x = "Years of Experience")

Salary_Data |> 
  group_by(`Education Level`) |> 
  summarize(n = n(), mean_salary = mean(Salary), std_dev = sd(Salary))

summary(Salary_Data)
```

### Conclusions
Using the Summary line and running > model <- lm(Salary ~ `Education Level`, data = Salary_Data)  summary(model) in the console, I found that the Education level was very meaningful when trying to determine the influence on variance in salary. With a 42.2% factor, this accounts for almost half of the R squared value. The PhD category showed a Calculated  Coefficient of 65,652 with a P value of .102, while the Masters showed a coefficient of 30,112 with a P value of .453. Both of these confirm our R squared findings that the education level is in fact statistically significant in regards to salary. 

> Below, interpret the results of this analysis (that is, describe what the results mean for your data and your hypotheses). Please also discuss any practical implications of these results (that is, explain how people might wish to think or act differently based on these results). [1 point]

Based on the data in the first graph, someone might assume that simply "waiting it out" could result in ending up with a higher salary. It is impractical to assume that age alone will result in a higher salary. If someone works until they are 50 in one industry, and then switches industries, they should not automatically assume they are going to receive the same or a similar salary. Instead, their salary will be based on experience, certifications, tenure, locale, and various other factors. This graph could very well be used to support a desire for higher education, especially noting that before the age of 50 there is a severe difference in the average salaries between the education levels. The conclusion someone could draw from this is that combining education with experience, age and many other factors, will result in the highest salary possible. This means the alternative hypothesis is true. 


# Submitting Project 3

After you've completed all parts of this assignment to your satisfaction, click on the `Knit` button in the toolbar at the top of this pane. When/if prompted, please indicate that you wish to knit to an HTML file. Knitting to PDF or Word are fancy tricks but take some more setting up, and that's beyond the scope of this class. 

After RStudio has finished knitting the file, it will bring it up in a special RStudio interface. **You should use this interface to review the document and make sure that everything appears as you want.** Then, however, you should navigate to the same folder where you have stored this `.Rmd` file, and you will find a `.html` file with the same name in that same folder. You can open that file to view this in your browser; more importantly, this is the file that you should submit to Canvas to complete this assignment.

I cannot accept any submission of your work besides the knit `.html` file! If you are having trouble knitting, please get in touch with me instead of submitting the `.Rmd` file and hoping it will work instead. Likewise, please don't attempt to manually convert this `.Rmd` file into an `.html` file—that will not work either.

Speaking of having your data, though, it would be nice to have access to all the work you've done in addition to the `.html` file. Please open GitHub desktop, navigate to your repository for this class, *commit* the changes that you've made in completing this project, and *push* those changes to GitHub. I will also ask you to do this if you need any help along the way.